Symmetric Peer-to-Peer Applications

### 2022-08-25 tpd-22.tex
...

The main job of the middleware is to deal with the uncertainties of the
network in order to preserve the symmetric behavior across peers.
More specifically, the middleware ensures that all peers
    (a) meet event deadlines,
    (b) advance in sync in real time, and
    (c) can leave and join the network and remain symmetric:
%
\begin{enumerate}
\item \textbf{Event Deadlines:}
Due to the inherent latency of networks, peers generate events with a
deliberate delay so that they reach other peers in time to be applied in sync
in a common deadline.
%We first assume a predetermined delay for the sake of simplicity.
However, a peer that is locally ahead of time may receive an event that should
have been applied in the past, event considering the delay.
In this case, the middleware automatically rolls back the peer, applies the
event at the correct time, and then fast forwards the peer, re-applying all
remaining events up to the real time.
%
\item \textbf{Time Synchronization:}
Since peers run in different machines, their timelines will inevitably be out
of sync because:
    (a) applications are launched locally at different times, and
    (b) internal clocks may diverge over time.
The middleware assumes the maximum time among all peers the be the
``correct real time'', so that other peers must fast forward to catch up with
it.
Therefore, in order to determine the maximum time and synchronize the clocks,
the middleware proceeds as follows:
    (a) peers broadcast \code{SYNC} events periodically with their local
        times, and
    (b) if a peer is behind a received \code{SYNC}, it advances its frames
        proportionally faster to catch up in time.
%
\item \textbf{Peer Churn:}
Considering that we target peer-to-peer networks, it is important that the
middleware can handle arrival and departure of peers seamlessly.
When a peer leaves the network, nothing needs to be done.
The middleware just needs to ensure that a peer that joins the network receives
all events in order, which is trivial since they all hold an absolute
timestamp.
\end{enumerate}

@@

{\footnotesize
\begin{verbatim}
int main (void) {
    p2p_loop (
        1,              // peer identifier
        50,             // simulation FPS
        sizeof(G), &G,  // pre-allocated full state
        cb_ini,         // init/quit callback
        cb_sim,         // simulation callback
        cb_ren,         // rendering callback
        cb_evt          // event triggering callback
    );
}
\end{verbatim}
}

The middleware assumes that each peer is assigned a unique identifier in a
contiguous range.
The FPS must be the same in all peers to ensure that the simulation executes
exactly the same.
The global variable \code{G} must hold the full application state.
If dynamic allocation is required, it is necessary to hold a finite memory
pool in \code{G} with a custom allocator.
For instance, the moving rectangle application only requires to hold the
current position and direction:

@@

@@

The middleware calls \code{cb\_sim} on every frame to affect the simulation
state of \code{G} deterministically.
The callback must never perform any side effects, such as stateful calls or
rendering the frame:
The callback receives the occurring input event if any.
There are two predetermined events:
    \code{P2P\_EVT\_INIT} called at the beginning of the simulation, and
    \code{P2P\_EVT\_TICK} called at every frame.
The application-specific events must be declared by the programmer in an
enumeration.
In our application, we define a new event \code{P2P\_EVT\_KEY} to represent key
presses:

Our callback \code{cb\_sim}
    (a) resets the rectangle position and speed on \code{INIT},
    (b) increments the position based on the current speed on \code{TICK}, and
    (c) modifies the speed based on \code{KEY}:

{\footnotesize
\begin{verbatim}
void cb_sim (p2p_evt evt) {
    switch (evt.id) {
        case P2P_EVT_INIT: // resets pos and speed
            G.x = G.y = G.dx = G.dy = 0;
            break;
        case P2P_EVT_TICK:  // increases position
            G.x += G.dx * VEL;
            G.y += G.dy * VEL;
            break;
        case P2P_EVT_KEY:  // modifies speed
            switch (evt.pay.i1) {
                case SDLK_UP:
                    G.dx = 0;
                    G.dy = -1;
                    break;
                <...> // same for other keys
            }
            break;
    }
}
\end{verbatim}
}

The calls to \code{cb\_sim} normally happen during real-time simulation, i.e.,
while the user is interacting with the application.
However, if the peer is out of sync, the middleware may ``time travel'' and
call the simulation many times to go back and forth and catch up with real
time.
This is the reason why we define \code{INIT} as an event: time travelling
requires to simulate, since the

This is the reason why \code{cb\_sim} should be split from the rendering
callback \code{cb\_ren}~\cite{tml.js}.
%, which the middleware calls only on real-time frames.
In our example, this callback just needs to redraw the rectangle at the
current position:

@@

Finally, the \code{cb\_evt} callback is called by the middleware in real time,
whenever a local SDL event occurs.
This callback has two goals:
    (a) broadcast application events, and
    (b) provide instant feedback to the user.
Not all local low-level events may generate application events to be broadcast
in the network.
The callback is allowed to perform side effects, such as modifying the network
topology, or exhibiting some feedback on the screen.
In our example, we only generate application events for the 5 key events, and
we also exhibit the pressed key in real time:

>>>>

The main job of the middleware is to deal with the uncertainties of the
network in order to preserve the symmetric behavior across peers.
More specifically, the middleware ensures that all peers
    (a) meet event deadlines,
    (b) advance in sync in real time, and
    (c) can leave and join the network and remain symmetric:
%
\begin{enumerate}
\item \textbf{Event Deadlines:}
Due to the inherent latency of networks, peers generate events with a
deliberate delay so that they reach other peers in time to be applied in sync
in a common deadline.
%We first assume a predetermined delay for the sake of simplicity.
However, a peer that is locally ahead of time may receive an event that should
have been applied in the past, event considering the delay.
In this case, the middleware automatically rolls back the peer, applies the
event at the correct time, and then fast forwards the peer, re-applying all
remaining events up to the real time.
As we detail further, rolling back a peer requires to simulate its whole
execution from the beginning (or from a snapshot in between).
%
\item \textbf{Time Synchronization:}
Since peers run in different machines, their timelines will inevitably be out
of sync because:
    (a) applications are launched locally at different times, and
    (b) internal clocks may diverge over time.
The middleware assumes the maximum time among all peers the be the
``correct real time'', so that other peers must fast forward to catch up with
it.
Therefore, in order to determine the maximum time and synchronize the clocks,
the middleware proceeds as follows:
    (a) peers broadcast \code{SYNC} events periodically with their local
        times, and
    (b) if a peer is behind a received \code{SYNC}, it advances its frames
        proportionally faster to catch up in time.
%
\item \textbf{Peer Churn:}
Considering that we target peer-to-peer networks, it is important that the
middleware can handle arrival and departure of peers seamlessly.
When a peer leaves the network, nothing needs to be done.
The middleware just needs to ensure that a peer that joins the network receives
all events in order, which is trivial since they all hold an absolute
timestamp.
\end{enumerate}

@@

{\footnotesize
\begin{verbatim}
int main (void) {
    p2p_loop (
        1,              // peer identifier
        50,             // simulation FPS
        sizeof(G), &G,  // pre-allocated full state
        cb_ini,         // init/quit callback
        cb_sim,         // simulation callback
        cb_ren,         // rendering callback
        cb_evt          // events callback
    );
}
\end{verbatim}
}

The middleware assumes that each peer is assigned a unique identifier in a
contiguous range.
The FPS must be the same in all peers to ensure that the simulation executes
exactly the same.

\subsubsection{Application State}

The global variable \code{G} must hold the full application state.
If dynamic allocation is required, it is necessary to hold a finite memory
pool in \code{G} with a custom allocator.
For instance, the moving rectangle application only requires to hold the
current position and direction:

@@

\subsubsection{Application Events}

The application generates events that need to be broadcast to the network so
that all peers behave the same.
In this sense, application events may differ from low-level local system
events, which may or may not be mapped 1-to-1 in the callback \code{cb\_evt}.

There middleware predefines two events for all applications:
    \code{P2P\_EVT\_START} represents the beginning of the simulation, and
    \code{P2P\_EVT\_TICK} called at every frame.
The application-specific events must be declared by the programmer in an
enumeration.
In our application, we define a new event \code{P2P\_EVT\_KEY} to represent key
presses:

\subsubsection{Initialization Callback}

@@

\subsubsection{Simulation Callback}

The middleware calls \code{cb\_sim} on every frame, passing an application
event, to affect the simulation state \code{G} deterministically.
The callback must never perform any side effects, such as stateful calls or
rendering frames.
In our example, we modify the state of the rectangle as follows:
    (a) reset its position and speed on \code{START},
    (b) increment its position based on the current speed on \code{TICK}, and
    (c) modify its speed based on \code{KEY}:

{\footnotesize
\begin{verbatim}
void cb_sim (p2p_evt evt) {
    switch (evt.id) {
        case P2P_EVT_START: // resets pos and speed
            G.x = G.y = G.dx = G.dy = 0;
            break;
        case P2P_EVT_TICK:  // increases position
            G.x += G.dx * VEL;
            G.y += G.dy * VEL;
            break;
        case P2P_EVT_KEY:  // modifies speed
            switch (evt.pay.i1) {
                case SDLK_UP:
                    G.dx = 0;
                    G.dy = -1;
                    break;
                <...> // same for other keys
            }
            break;
    }
}
\end{verbatim}
}

The calls to \code{cb\_sim} normally happen during real-time simulation, i.e.,
while the user is interacting with the application.
However, if the peer is out of sync, the middleware may ``time travel'' and
call the simulation many times to go back and forth and catch up with real
time.
This is the reason why we define \code{START} as an event like any other, since
travelling requires to simulate the execution from the beginning, event by
event.

\subsubsection{Rendering Callback}

Time travelling is also the reason why \code{cb\_sim} must be split from the
rendering callback \code{cb\_ren}~\cite{tml.js}, which would otherwise render
the screen multiple times while travelling.
%, which the middleware calls only on real-time frames.
In our example, \code{cb\_ren} just needs to redraw the rectangle at the
current position:

@@

\subsubsection{Events Callback}

The \code{cb\_evt} callback is called by the middleware in real time, whenever
a local SDL event occurs.
This callback has two goals:
    (a) map and broadcast application events, and
    (b) provide instant feedback to the user.
As mentioned previously, not all local low-level events may generate
application events to be broadcast in the network.
The callback is allowed to perform side effects, such as modifying the network
topology, or exhibiting some feedback on the screen.
In our example, we only generate application events for the 5 key events, and
we also exhibit the pressed key in real time:
